### 2024-02-02 paper.tex
final touches

\begin{figure*}
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
	\end{center}
	\caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with \new{conventional methods based on genome tiling (B) or our proposed methods (C)}.
		\new{
		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
		Either VMRs, ENCODE regulatory elements, 100~kb genomic tiles or promoter regions (TSS\textpm2kb) were subjected to iterative PCA and UMAP.
		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
	    }
		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
		\new{
		\textbf{(F)} Proportion of VMRs (blue dot) that have at least 1~bp overlap with other genomic features.
		The black range illustrates the minimum and maximum overlap observed for 100 permutations of randomly shuffled VMRs.
		}
	}
	\label{fig:score}
\end{figure*}

\section{Finding differentially methylated regions}

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
	\end{center}
	\caption{\small \textbf{Identification of differentially methylated regions.}\\
		The chromosome is divided into windows overlapping by a small and fixed size.
		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
	\label{fig:dmrscan}
\end{figure*}

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
	    \includegraphics[width=\textwidth]{figures/SFig_benchmark.png}
	\end{center}
	\captionof{figure}{\small \new{ \textbf{Benchmark of several combinations of analysis methods on four single-cell methylation data sets.}\\
	\textbf{(A-E)} Benchmarking the ability to separate groups of cells based on CpG methylation data, using different analysis approaches.
	Methylation matrices were obtained either by averaging CpG methylation in genomic intervals (dotted lines, hollow points) or using the shrunken mean of residuals (solid lines and points).
	Each analysis was performed for the full data sets and sub-samples of each.
	Four sets of genomic intervals (VMRs, ENCODE regulatory elements, 100~kb tiles or promoters) were quantified and separately subjected to four dimensionality reduction techniques (see Methods for details):
	iterative PCA as proposed by us, mean-imputed PCA on intervals with high sequencing coverage \citep{luo2017single}, mean-imputed PCA on all intervals, and MOFA+ \citep{argelaguet2020mofa} on all intervals.
	For each combination of methods, we quantified the ability to use distance in 15-dimensional PCA space to separate ground-truth cell labels reported by the authors (neighbor score).
	\textbf{(A)} Separation of neural cell types/states annotated based on scRNA-seq of the same cells \citep{kremer_scnmt}.
	\textbf{(B)} Separation of neuron sub-types, annotated by averaging CH-methylation (mCH) in 100~kb genomic tiles \citep{luo2017single}.
	\textbf{(C)} Separation of human colorectal cancer sampling regions (primary tumor, normal adjacent tissue, lymph node metastasis, liver metastasis before and after treatment, omental metastasis) \citep{bian2018single}.
	\textbf{(D-E)} Separation of three properties of mouse embryo cells:
	The developmental stage (E4.5, E5.5, E6.5, E7.5), the germ layer or extra-embryonic tissue of origin, or the cell lineage.
	Germ layer and lineage (\textbf{E}, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
	}}
	\label{fig:score_luo}

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
		\includegraphics[width=.85\textwidth]{figures/SFig_benchmark_resources.png}
	\end{center}
	\captionof{figure}{\small \new{\textbf{VMRs are as informative as regulatory regions, but more resource-efficient.}\\
	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citep{kremer_scnmt} and \citet{luo2017single}.
	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.
	\textbf{(C-D)} Runtime (top) and RAM usage (bottom) of the two dimensionality reduction techniques shown in A and B.
	}}
	\label{fig:resource_usage}
\end{minipage}

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
		\includegraphics[width=.85\textwidth]{figures/SFig_100k_and_CH.png}
	\end{center}
	\captionof{figure}{\small \new{\textbf{\texttt{scbs} performs well on CH-methylation data and on extremely large data sets.}\\
    \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
	top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single};
	center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs;
	bottom: shrunken mean of residuals to quantify CpG methylation in CpG-VMRs.
	\textbf{(B)} UMAP obtained when analysing 100\,350 neural single-cell methylomes \citep{liu2021dna} with the default \texttt{scbs} workflow.
	The depicted cell labels were reported by \citet{liu2021dna}.
	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which needs to be run only once per data set. The speed of this step can be increased considerably by processing chromosomes in parallel (each takes 6 to 12 hours depending on its size).}
	}}
	\label{fig:umaps}
\end{minipage}

>>>>

\begin{figure*}
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
	\end{center}
	\caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with \new{conventional methods based on genome tiling (B) or our proposed methods (C)}.
		\new{
		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
		Either VMRs, ENCODE regulatory elements, 100~kb genomic tiles or promoter regions (TSS\textpm2kb) were subjected to iterative PCA and UMAP.
		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
	    }
		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
		\new{
		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
		The black range illustrates the minimum and maximum overlap observed for 100 permutations of randomly shuffled VMRs.
		}
	}
	\label{fig:score}
\end{figure*}

\section{Finding differentially methylated regions (DMRs)}

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
	\end{center}
	\caption{\small \textbf{Identification of differentially methylated regions (DMRs).}\\
		The chromosome is divided into windows overlapping by a small and fixed size.
		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
	\label{fig:dmrscan}
\end{figure*}

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
	    \includegraphics[width=\textwidth]{figures/SFig_benchmark.png}
	\end{center}
	\captionof{figure}{\small \new{ \textbf{Benchmark of several combinations of analysis methods on four single-cell methylation data sets.}\\
	\textbf{(A-E)} Benchmarking the ability to separate groups of cells based on CpG methylation data, using different analysis approaches.
	Methylation matrices were obtained either by averaging CpG methylation in genomic intervals (dotted lines, hollow points) or using the shrunken mean of residuals (solid lines and points).
	Each analysis was performed for the full data sets and sub-samples of each.
	Four sets of genomic intervals (VMRs, ENCODE regulatory elements, 100~kb tiles or promoters) were quantified and separately subjected to four dimensionality reduction techniques (see Methods for details):
	iterative PCA as proposed by us, mean-imputed PCA on intervals with high sequencing coverage \citep{luo2017single}, mean-imputed PCA on all intervals, and MOFA+ \citep{argelaguet2020mofa} on all intervals.
	For each combination of methods, we quantified the ability to use distance in 15-dimensional PCA space to separate ground-truth cell labels reported by the authors (neighbor score).
	\textbf{(A)} Separation of neural cell types/states annotated based on scRNA-seq of the same cells \citep{kremer_scnmt}.
	\textbf{(B)} Separation of neuron sub-types, annotated by averaging CH-methylation (mCH) in 100~kb genomic tiles \citep{luo2017single}.
	\textbf{(C)} Separation of human colorectal cancer sampling regions (primary tumor, normal adjacent tissue, lymph node metastasis, liver metastasis before and after treatment, omental metastasis) \citep{bian2018single}.
	\textbf{(D-E)} Separation of three properties of mouse embryo cells:
	The developmental stage (E4.5, E5.5, E6.5, E7.5), the germ layer or extra-embryonic tissue of origin, or the cell lineage.
	Germ layer and lineage (E, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
	}}
	\label{fig:score_luo}

@@

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
		\includegraphics[width=.85\textwidth]{figures/SFig_benchmark_resources.png}
	\end{center}
	\captionof{figure}{\small \new{\textbf{VMRs are more informative than an equal number of regulatory regions with high sequencing coverage.}\\
	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citet{kremer_scnmt} and \citet{luo2017single}.
	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.
	\textbf{(C-D)} Runtime (top) and RAM usage (bottom) of the two dimensionality reduction techniques shown in A and B.
	}}
	\label{fig:resource_usage}
\end{minipage}

\noindent%
\begin{minipage}{\linewidth}
	\begin{center}
		\includegraphics[width=.85\textwidth]{figures/SFig_100k_and_CH.png}
	\end{center}
	\captionof{figure}{\small \new{\textbf{\texttt{scbs} performs well on CH-methylation data and on extremely large data sets.}\\
    \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
	top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single};
	center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs;
	bottom: shrunken mean of residuals to quantify CpG methylation in CpG-VMRs.
	\textbf{(B)} UMAP obtained when analysing 100\,350 neural single-cell methylomes \citep{liu2021dna} with the default \texttt{scbs} workflow.
	The depicted cell labels were reported by \citet{liu2021dna}.
	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which is run only once per data set. The speed of this step can be increased considerably by processing chromosomes in parallel, which reduced the total runtime to approximately two days in our case.}
	}}
	\label{fig:umaps}
\end{minipage}
